Anomaly detection with kernel density estimation on manifolds

### 2022-04-27 paper/paper.Rmd
Add example of S1 in appendix

#### Example

Manifold: S2

$\theta \sim U(0, \pi/2)$

$X = \cos(\theta)$

$$

>>>>

#### Example: $S^1$

Manifold: $S^1$ with center 0 and radius $r=1$

Polar coordinates: $(r, \theta)$

Uniformly distributed angle: $\theta \sim U(0, \pi/2)$ with $f(\theta) = 2 / \pi$ for $\theta \in [0, \pi /2]$.

Cartesian coordinates: $(x, y)$ where $x = \cos(\theta) \in [0, 1]$ and $y = \sin(\theta) \in [0, 1]$

Density of $\theta$: $f(\theta) = 2 / \pi$

Density of $x$: $f(x) = \frac{2}{\pi \sqrt{1-x^2}}$

The CDF of X is given by
$$
P(X \leq x) = P(\cos(\theta) \leq x) \\
= P(\theta \geq \arccos(x)) \\
= \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
=  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
= 1 - \frac{2}{\pi} \arccos(x).
$$
Then the density of X is the derivative of the CDF w.r.t. $x$
$$
f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
= \frac{2}{\pi \sqrt{1 - x^2}}.
$$
Given $(x, y) \in \mathbb{R}^2$ as sample points from $S^1$, we could estimate the density of each sample point $X_i \in S^1, i = 1, \dots, N$ and compare it with its true density $f(x)$.

<!-- we could reduce the dimension to $d = 1$ and estimate the density of the 1-dimensional embedding$. -->

<!-- First, we use the kernel density estimator in \@ref(equ:kde) with a fixed bandwidth using the plug-in bandwidth selector in the R package *ks*.  -->

We use the density estimator in @Pelletier2005-vu. 
<!-- \begin{equation} -->
<!-- \label{eq:pelletier} -->
$$
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{h^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{h}),
$$
<!-- \end{equation} -->
where $d=2$, $h$ is the bandwidth, $\theta_{X_i}(p)$ is the volume density function at sample point $X_i$, and $d_g(p, X_i)$ is the geodesic distance between $p$ and $X_i$ along the manifold $S^1$.

Since the manifold is known ($S^1$), we know the geodesic distance is the arc length between two points on the manifold, which is given by $L(\gamma) = r(\beta - \alpha)$. Therefore, the geodesic distance is
$$
d_g(p, X_i) = \arccos(p) - \arccos(X_i)
$$
As stated in @Pelletier2005-vu, in terms of *geodesic normal coordinates* at $X_i$, the volume density function $\theta_{X_i}(p)$ equals the determinant of the metric $g$ expressed in these coordinates at the logarithm map $\exp_{X_i}^{-1}(p)$, i.e. $|g(\exp_p^{-1}(x_i))|$.

[When using normal coordinates(orthogonal basis) for tangent vector in $T_{X_i}S^1$, the exponential map has *geodesic normal coordinates*.]

The Riemannian metric $g$ for $S^1$ is given by the inner products on the tangent spaces of $S^1$. For each $X_i \in M$,
$$
g_{X_i}: T_{X_i}M \times T_{X_i}M \rightarrow \pmb{R},
$$
is smooth in a local neightborhood $U$ of $M$.

Then the volume density function $\theta_{X_i}(p)$ is the inner product of the tangent vectors produced by the logarithm map from $X_i$ to $p$, which is given by
$$
\theta_{X_i}(p) = \frac{1}{\sqrt{1 - X_i^2}}.
$$
Therefore, the estimator for $p \in S^1$ is given by
$$
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} |h|
^{-d} \frac{1}{\sqrt{1 - X_i^2}} K(|\arccos(p) - \arccos(X_i)|^{T}h^{-1}|\arccos(p) - \arccos(X_i)|).
$$
The bandwidth $h$ is selected by the plug-in bandwidth selector implemented by the R package *ks*.
